```{r}
library(dplyr)
library(ggplot2)
```

```{r}
activity <- read.csv(unz("activity.zip", "activity.csv"), colClasses = list(date = "Date"))

# Convert the intervals column into a factor, 
# so that during plotting/processing we won't treat them as integer numbers.
# (ie. we make the distance between 1155 and 1200 equal to the distance between 1145 and 1150)
# Also make the interval labels a bit more human-readable.
activity$interval <- factor(sprintf("%04d", activity$interval))
```

### Missing values / imputation

* We either have all intervals for a day or none
* There are 8 days missing in total
    * Two of them are next to each other so it might be dependent?

```{r}
ggplot(summarize(group_by(activity, date), miss_count = sum(is.na(steps))), aes(date, miss_count)) + geom_bar(stat = "identity")
summarize(group_by(activity, date), miss_count = sum(is.na(steps)))$miss_count
```

```{r}
summarize(group_by(activity, interval), miss_count = sum(is.na(steps)))$miss_count
```

#### Are close to zeroes a miss too?

```{r}
ggplot(summarize(group_by(activity, date), s = sum(steps)), aes(date, s)) + geom_line(group = 1)
```

```{r}
sfrm <- summarize(group_by(activity, date), S = sum(steps, na.rm = T), missing = sum(is.na(steps)) == 288)
ggplot(sfrm, aes(date, S)) + geom_line() + geom_point(aes(color = missing), size = 3) + scale_color_manual(values = c("black", "orange"))
```

#### Are misses/lows timeseries-dependent?

```{r}
ggplot(sfrm, aes(x = S, y = lag(S, 1), color = missing)) + geom_point(size = 3) + scale_color_manual(values = c("black", "orange")) + stat_smooth(method = lm)
```

```{r}
ggplot(sfrm, aes(x = S, y = lag(S, 1))) + geom_point(size = 3) + scale_color_manual(values = c("black", "orange")) + stat_smooth(method = lm)
```

#### How does the time series look like if we impute the average day

```{r}
sort(sfrm$S)
```

```{r}
sfrm$low.or.0 <- sfrm$S <= 126
M <- mean(sfrm$S[!sfrm$low.or.0])
sfrm$imp <- ifelse(sfrm$low.or.0, M, sfrm$S)
```

```{r}
ggplot(sfrm, aes(date, imp)) + geom_line() + geom_point(aes(color = missing), size = 3) + scale_color_manual(values = c("black", "orange"))
```

Interploation would we nicer maybe, but oh well...

#### Are missings and zeros/lows consistent on the interval-series level?

```{r}
plot(activity$step)
```

```{r}
activity$timestamp <- as.POSIXct(strptime(paste(activity$date, activity$interval), format = "%Y-%m-%d %H%M", tz = "UTC"))
ggplot(activity, aes(timestamp, steps)) + geom_line()
```

```{r fig.width = 16, fig.height = 8}
arrange(filter(sfrm, low.or.0), date)
frm <- filter(activity, date %in% as.Date(c("2012-10-01", "2012-10-02", "2012-10-03")))
ggplot(frm, aes(timestamp, steps)) + geom_line() + geom_point()

frm <- filter(activity, date %in% as.Date(c("2012-10-07", "2012-10-08", "2012-10-09")))
ggplot(frm, aes(timestamp, steps)) + geom_line() + geom_point()

frm <- filter(activity, date %in% as.Date(c("2012-10-31", "2012-11-01", "2012-11-02", "2012-11-03", "2012-11-04", "2012-11-05")))
ggplot(frm, aes(timestamp, steps)) + geom_line() + geom_point()

frm <- filter(activity, date %in% as.Date(c("2012-11-08", "2012-11-09", "2012-11-10", "2012-11-11", "2012-11-12")))
ggplot(frm, aes(timestamp, steps)) + geom_line() + geom_point()

frm <- filter(activity, date %in% as.Date(c("2012-11-13", "2012-11-14", "2012-11-15", "2012-11-16")))
ggplot(frm, aes(timestamp, steps)) + geom_line() + geom_point()


frm <- filter(activity, date %in% as.Date(c("2012-11-28", "2012-11-29", "2012-11-30")))
ggplot(frm, aes(timestamp, steps)) + geom_line() + geom_point()
```

